Staged Abstract Interpreters

### 2019-09-04 paper/50_stageabs.tex
revision

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter, now we begin describing the implementation of their
confluence -- a staged abstract interpreter.
Unsurprisingly, the staged abstract interpreter in this section has the same
abstract semantics with the unstaged version from Section~\ref{unstaged_abs}.
The approach from unstaged to staged abstract interpreter is modular,
and does not sacrifice soundness or precision. The designer of the
analyzer therefore has no need to rewrite the analysis. We first present the
staged lattices and staged monads, and then the staged version of primitive
operations, especially @close@, @ap_clo@ and @fix@. At last, we discuss several
optimizations and summarize our approach.

@@

In Section \ref{stagedpoly_lat}, we exploited the higher-kinded type @R@ to
achieve stage polymorphism. Now we instantiate the type @R@ to @Rep@ and
still use power sets as an example to present its staged version.
\begin{lstlisting}[escapechar=!]
  trait RepLattice[S] extends Lattice[S, Rep]
  def RepSetLattice[T]: RepLattice[Set[T]] = new RepLattice[Set[T]] {
    lazy val ⊥: !\hl{Rep[Set[T]]}! = Set[T]()
    lazy val ⊤: !\hl{Rep[Set[T]]}! = error("No representation for ⊤")
    def ⊑(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Boolean]}! = l1 subsetOf  l2
    def ⊔(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 union     l2
    def ⊓(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 intersect l2
  }
\end{lstlisting}

The trait @RepLattice@ is obtained by specializing @Lattice@ with @Rep@.
The @RepSetLattice@ is an instance of @RepLattice@, where the lattice operations
eventually are delegated to operations on a staged set data type, including
@subsetOf@, @union@ and @intersect@. We even do not have to change the
implementation code, but just the types -- from @Set[T]@ to @Rep[Set[T]]@, shown
in light gray code. The LMS library provides an implicit conversion that lifts a
current-stage set value to next-stage values. In the code generation part, these
operations emit their corresponding next-stage code. Again, other lattice
structures such as products and maps are implemented in a similar way.

@@

Now we can implement the staged abstract semantics, where @R@ is instantiated as
@Rep@, and the abstract components are reused from the unstaged version.

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure from the unstaged abstract interpreter,
but replacing \textit{all} the transformers to their staged version. The
following code shows this change. For readability, we squeeze the three inner
transformers (@IdM@ omitted) into a single monad @RepSetReaderStateM[A, B, C]@,
where @A@ is the type parameter for reader effect and @B@ for the state effects.
In our monad stack scheme, they are both instantiated with @Cache@. Now the
result type is a pair of two staged value: @Rep[Set[(Value, Store)]]@ and
@Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

@@

\paragraph{Primitive Operations} When deriving the staged concrete interpreter
from the unstaged counterpart, we notice that recursive calls of @ev@ on the
lambda body are shifted from @ap_clo@ to @close@. In other word, we need to
eagerly specialize the interpreter and generate code every time when we see a
$\lambda$ term, instead of lazily when we actually need to apply a function.
With no surprise, it is similar in the staged abstract interpreter,
additionally, we need to handle nondeterminism. So we keep our track and focus
on discussing closure representations and function applications for the staged
version.

@@

Our staging schema works by unfolding the interpreter over the input AST. In
practice, however, it would suffer from code explosion when analyzing
(specializing) large programs, which increases the compile time. If we generate
next stage programs running on JVM, such large generated program would also
cause runtime GC overhead. In this section, we present several optimizations
that largely mitigate these issues. For all that, implementing these
optimizations do not need to change the generic interpreter.

@@

\section{Discussion}

>>>>

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter, now we begin describing the implementation of
their confluence -- a staged abstract interpreter.  Unsurprisingly, the staged
abstract interpreter in this section has the same abstract semantics with the
unstaged version from Section~\ref{unstaged_abs}.  The approach that refactors
the unstaged one to the staged abstract interpreter is modular, and does not
sacrifice soundness or precision. The designer of the analyzer therefore has no
need to rewrite the analysis. We first present the staged lattices and staged
monads, and then the staged version of primitive operations, especially
@close@, @ap_clo@ and @fix@. At last, we discuss several optimizations.

@@

In Section \ref{stagedpoly_lat}, we exploited the higher-kinded type @R@ to
achieve stage polymorphism. Now we instantiate the type @R@ to @Rep@ and
still use power-set as an example to describe its staged version.
\begin{lstlisting}[escapechar=!]
  trait RepLattice[S] extends Lattice[S, Rep]
  def RepSetLattice[T]: RepLattice[Set[T]] = new RepLattice[Set[T]] {
    lazy val ⊥: !\hl{Rep[Set[T]]}! = Set[T]()
    lazy val ⊤: !\hl{Rep[Set[T]]}! = error("No representation for ⊤")
    def ⊑(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Boolean]}! = l1 subsetOf  l2
    def ⊔(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 union     l2
    def ⊓(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 intersect l2
  }
\end{lstlisting}

The trait @RepLattice@ is obtained by specializing @Lattice@ with @Rep@.
The @RepSetLattice@ is an instance of @RepLattice@, where the lattice operations
are eventually delegated to operations on a staged set data type, such as
@subsetOf@, @union@ and @intersect@. We even do not have to change the
implementation code, except the types -- from @Set[T]@ to @Rep[Set[T]]@, as shown
in light gray code. The LMS library provides implicit conversions that lift
current-stage values to next-stage values. In the code generation part, these
operations emit their corresponding next-stage code. Again, other lattice
structures such as products and maps are implemented in a similar way.

@@

Now we can implement the staged abstract semantics, where @R@ is instantiated as
@Rep@, and the types of abstract components are reused from the unstaged version.

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure from the unstaged abstract interpreter,
but replacing \textit{all} the transformers to their staged version. The
following code shows this change. We manually squeeze the three inner
transformers (@IdM@ omitted) into a single monad @RepSetReaderStateM[A, B, C]@,
where @A@ is the type parameter for the reader effect and @B@ for the state effect.
In our monad stack scheme, they are both instantiated with type @Cache@. Now the
grounded result type is a pair of two staged value: @Rep[Set[(Value, Store)]]@ and
@Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

@@

\paragraph{Primitive Operations} When deriving the staged concrete interpreter
from the unstaged counterpart, we notice that the recursive calls of @ev@ on the
lambda body are shifted from @ap_clo@ to @close@. In other words, we need to
eagerly specialize the interpreter and generate code every time when we see a
$\lambda$ term, instead of lazily when applications happen.
With no surprise, it is similar in the staged abstract interpreter;
additionally, we have to handle nondeterminism occurred by over-approximation
of runtime behaviors.
In the rest of this part, we keep our roadmap and focus on discussing closure
representations and function applications for the staged version.

@@

Our staging schema works by unfolding the interpreter over the input AST. In
practice, however, it would suffer from code explosion when analyzing
(specializing) large programs, which increases the compile time. If we generate
next stage programs running on JVM, such large generated program would also
potentially cause runtime GC overhead. In this section, we present several
optimizations that largely mitigate these issues. For all that, implementing
these optimizations do not need to change the generic interpreter.

@@

\section{Discussion} \label{discussion}
